---
title: "10X ik"
author: "Koen Van den Berge"
date: "11/7/2019"
output: html_document
---

```{r}
library(Matrix)
library(aroma.light)
countsFull <- readMM("~/Downloads/GSE128889_RAW/GSM3717977_SCmurinep12_matrix.mtx.gz")
ft <- read.table("~/Downloads/GSE128889_RAW/GSM3717977_SCmurinep12_genes.tsv.gz")
bc <- read.table("~/Downloads/GSE128889_RAW/GSM3717977_SCmurinep12_barcodes.tsv.gz")
rownames(countsFull) <- as.character(ft[,2])
```

# Preprocessing, filtering and trajectory inference

```{r}
keep <- rowSums(countsFull > 1) >= 400
counts <- countsFull[keep,]
rownames(counts) <- ft[keep,2]
countsFQ <- normalizeQuantileRank(as.matrix(counts))
rownames(countsFQ) <- ft[keep,2]

pca <- prcomp(t(log1p(countsFQ)), scale. = FALSE)
rd <- pca$x[,1:8]

plot(rd[,1:2], pch=16, cex=1/2)

plotByGene <- function(rd, geneCount, ng=10, main=NULL, ...){
  pal <- wesanderson::wes_palette("Zissou1", n=ng, type="continuous")
  gg <- Hmisc::cut2(geneCount, g=ng)
  plot(rd, pch=16, cex=1/2, col=pal[gg], main=main, ...)
}

# progenitors
plotByGene(rd[, 1:2], countsFQ["Dpp4",], ng=2, main="Dpp4")
plotByGene(rd[, 1:2], countsFQ["Wnt2",], ng=4, main="Wnt2")

# group 2 cells
plotByGene(rd[, 1:2], countsFQ["Icam1",], ng=4, main="Icam1")
plotByGene(rd[, 1:2], countsFQ["Dlk1",], ng=4, main="Pref1") #Pref1 is also called Dlk1

# group 3 cells
plotByGene(rd[, 1:2], countsFQ["Clec11a",], ng=4, main="Clec11a")

# group 4 cells
plotByGene(rd[, 1:2], countsFull["Wnt6",], ng=2, main="Wnt6")


# cluster
set.seed(9)
cl <- kmeans(rd, centers = 10)
color = grDevices::colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]
pal <- sample(color, 12)
plot(rd[,1:2], pch=16, cex=1/2, col=pal[cl$cluster])
legend("bottomleft", legend=1:10, pch=16, col=pal, bty='n')

# cluster 3 and 7 are progenitor cells
# cluster 1, 5 and 10 are group 2 cells
# cluster 2 and 6 are group 3 cells
keepCells <- which(cl$cluster %in% c(1,2,3,5,7,10))
countsFQ2 <- countsFQ[,keepCells]
rd2 <- prcomp(t(log1p(countsFQ2)), scale. = FALSE)

### umap
library(umap)
rdUmap <- umap(rd2$x[,1:20])
plot(rdUmap$layout, pch=16, cex=1/3)

rafalib::mypar(mfrow=c(3,2))
# progenitors
plotByGene(rdUmap$layout, countsFQ2["Dpp4",], ng=2, main="Prog: Dpp4")
plotByGene(rdUmap$layout, countsFQ2["Wnt2",], ng=4, main="Prog: Wnt2")

# group 2 cells
plotByGene(rdUmap$layout, countsFQ2["Icam1",], ng=4, main="G2: Icam1")
plotByGene(rdUmap$layout, countsFQ2["Dlk1",], ng=4, main="G2: Pref1") #Pref1 is also called Dlk1

# group 3 cells
plotByGene(rdUmap$layout, countsFQ2["Clec11a",], ng=4, main="G3: Clec11a")

# group 4 cells
plotByGene(rdUmap$layout, countsFull["Wnt6",keepCells], ng=2, main="G4: Wnt6")

# cluster
set.seed(12)
pal <- wesanderson::wes_palette("Darjeeling1", n=6, type="continuous")
cl <- kmeans(rdUmap$layout, centers = 6)
plot(rdUmap$layout, pch=16, cex=1/2, col=pal[cl$cluster])
legend("topleft", legend=1:6, pch=16, col=pal[1:6], bty='n')

# slingshot
library(slingshot)
lin <- getLineages(rdUmap$layout, clusterLabels=cl$cluster, 
                   start.clus=1, end.clus=c(5,2))
plot(rdUmap$layout, pch=16, cex=1/2, col=pal[cl$cluster])
lines(lin, lwd=2)
crv <- getCurves(lin)
plot(rdUmap$layout, pch=16, cex=1/2, col=pal[cl$cluster],
     xlab="UMAP1", ylab="UMAP2", bty='l', asp=1)
lines(crv, lwd=2, col="black")
saveRDS(crv, file="crvUmap.rds")
```

# Evaluate optimal number of knots

8 knots seem appropriate.

```{r}
library(tradeSeq)
set.seed(3)
aicMat1 <- evaluateK(countsFQ2, k=3:10, nGenes=200, sds=crv)
set.seed(4)
aicMat2 <- evaluateK(countsFQ2, k=3:10, nGenes=200, sds=crv)
```

# Model fitting and inference

```{r}
sce <- fitGAM(countsFQ2, sds=crv, nknots=8)
saveRDS(sce, file="~/data/sceAdipose.rds")
```

# Progenitor cell markers other than Dpp4 and Wnt2 (startVsEndTest)

```{r}
seRes <- startVsEndTest(sce)
ose <- order(seRes$waldStat, decreasing=TRUE)
which(rownames(seRes)[ose] == "Wnt2") # ranked 263
which(rownames(seRes)[ose] == "Dpp4") # ranked 450

head(seRes[ose,], n=16)

rafalib::mypar(mfrow=c(3,2))
sapply(1:6, function(ii){
  gene <- rownames(seRes)[ose][ii]
  print(plotByGene(rdUmap$layout, countsFQ2[gene,], ng=4, main=gene, asp=1))
})
sapply(7:12, function(ii){
  gene <- rownames(seRes)[ose][ii]
  print(plotByGene(rdUmap$layout, countsFQ2[gene,], ng=4, main=gene, asp=1))
})


#figure for paper
rafalib::mypar(mfrow=c(3,2))
genes <- c("Dpp4", "Wnt2", 
           "Pi16", "Akr1c18",
           "Fn1", "Fbn1")
sapply(genes, function(gene){
  print(plotByGene(rdUmap$layout, countsFQ2[gene,], ng=4, main=gene, bty='n',
                   xlab="UMAP1", ylab="UMAP2"))
})
```


# Differentiated group 2 and group 3 markers (diffEndTest)

Note that this is only restricted to the end points and is therefore limited.

```{r}
deRes <- diffEndTest(sce)
ode <- order(deRes$waldStat, decreasing=TRUE)


head(seRes[ode,], n=10)

rafalib::mypar(mfrow=c(3,2))
sapply(1:6, function(ii){
  gene <- rownames(deRes)[ode][ii]
  print(plotByGene(rdUmap$layout, countsFQ2[gene,], ng=4, main=gene, asp=1))
})
```

# earlyDETest for entire cluster of group 2 and 3 cells

```{r}
plotGeneCount(crv, countsFQ2, models=sce, clusters = cl$cluster)
resedt <- earlyDETest(sce, knots=c(3,7))
oedt <- order(resedt$waldStat, decreasing=TRUE)

head(resedt[oedt,], n=10)

rafalib::mypar(mfrow=c(3,2))
sapply(1:6, function(ii){
  gene <- rownames(resedt)[oedt][ii]
  print(plotByGene(rdUmap$layout, countsFQ2[gene,], ng=4, main=gene, asp=1))
})


#figure for paper
rafalib::mypar(mfrow=c(1,2))
genes <- c("Mgp", "Meox2")
sapply(genes, function(gene){
  print(plotByGene(rdUmap$layout, countsFQ2[gene,], ng=4, main=gene, bty='n',
                   xlab="UMAP1", ylab="UMAP2", asp=1))
})

#figure for paper
rafalib::mypar(mfrow=c(1,2))
genes <- c("H19", "Col14a1")
sapply(genes, function(gene){
  print(plotByGene(rdUmap$layout, countsFQ2[gene,], ng=4, main=gene, bty='n',
                   xlab="UMAP1", ylab="UMAP2", asp=1))
})
```

